---
title: "STAA57 Draft Notebook"
author: "Faiza Khan - 1010864398, Adyan Milki Student Number: "
date: "2026-03-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Description of the Data
Description of the variables and the data\
answer: will be given in the repo\

### Wages by education level:

YEAR: Numerical data, ranging from 2006 to 2024
GEOGRAPHY: Categorical type, Canada or Ontario
IMMIGRANT: Categorical type, The immigration statuses include: total landed immigrants (very recent immigrants, recent immigrants, established immigrants), non-landed immigrants and born in Canada.
TYPE.OF.WORK: Type of work includes full-time and part-time.
WAGE.RATE: Categorical, Median hourly wage, Wages include average weekly wage rate.
EDUCATION: Categorical type, The educational levels include: 0-8 yrs., some high school, high school graduate, some post-secondary, post-secondary certificate diploma and university degree.
AGE.GROUP: Categorical type, The age groups available in the dataset are: 15+, 25+, 25-34, 25-54 and 25-64.
Both.Sexes: Median wage rate of both sexes
Men: Median wage rate of men
Women: Median wage rate of Women



# Background of Data

A little background about the data (who collected it in what context)\
answer: will be given in the repo\

The data used in this report was collected and published by the Government of Ontario through the open data portal. It includes datasets related to wages by education level, labour force estimates by educational attainment, and employment characteristics of students. These datasets are designed to provide insight into economic and workforce trends across the province and are regularly updated to support public research and policy analysis.

For this project, we focus on data that captures the relationship between post-secondary education level and career outcomes. Key variables include average wages, employment and unemployment rates, and labour force participation across different levels of educational attainment (e.g., less than high school, high school diploma, college, and university). These variables include both quantitative data (such as wages and rates) and categorical data (such as education level), making them well-suited for analysis using statistical and data science methods.

By working with this real-world dataset, we aim to identify patterns and relationships between education and career success, compare outcomes across different education levels, and evaluate how strongly education level is associated with economic opportunities in Ontario.

This data is publicly available through Ontario’s open data catalogue and is licensed under the Open Government Licence Ontario.


# Research Question
What is the over all research question that you are trying to answer in your report\

How does post-secondary education level affect career outcomes and measures of career success in Ontario?\

We explore this through three key questions:

1. How does education level affect average wages in Ontario?
2. How does education level influence employment and unemployment rates?
3. How does education level affect participation in the labour force (i.e., likelihood of being employed or actively seeking work)?

Additionally, we aim to develop a predictive model to estimate career outcomes (such as wages and employment status) based on education level.

# Analysis portion


```{r}
# libraries 
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)

# importing datasets (in order of importance)
WBE = read.csv("~/STAA57-Project/datasets/Wages by education level.csv")

```

```{r}
# table 1, avg wage rate for adults over 25 based on education level

table_2024 <- WBE %>%
  filter(YEAR == 2024,
         GEOGRAPHY == "Canada", 
         AGE.GROUP == "25 +",
         TYPE.OF.WORK == "All employees"	
         ) %>%
  group_by(EDUCATION) %>%
  summarise(avg_wage = mean(Both.sexes, na.rm = TRUE))

table_2024 <- table_2024 %>%
  mutate(EDUCATION = trimws(EDUCATION))


table_2024 = table_2024 %>%
  mutate(EDUCATION = factor(EDUCATION,
    levels = c(
      "0 - 8 years",
      "Some high school",
      "High school graduate",
      "Some post-secondary",
      "Post-secondary certificate or diploma",
      "Bachelor's degree",
      "University degree",
      "Above bachelor's degree"
    )
  )) %>%
  arrange(EDUCATION) %>%
  filter(!is.na(EDUCATION))

table_2024

```



Few tables (appropriately labeled), Few graphs (appropriately labeled)\

Use of confidence interval and test of hypothesis\

Use of bootstrapping, Use of regression analysis (linear or nonlinear regression [week-9] or logistic regression[week-10] or decision tree/random forest [Week-11]). Interpretation of the regression parameters when applicable.\

Use of cross validation\

# Final Summary
A final summary of your findings\

In conclusion:



# Apendix 
An appendix which shows all your codes\
```{r}
# libraries 
library(tidyverse)
library(ggplot2)
library(lubridate)

# importing datasets (in order of importance)
WBE = read.csv("~/STAA57-Project/datasets/Wages by education level.csv")
LFE1519 = read.csv("~/STAA57-Project/datasets/Labour force estimates by education-2015-2019.csv")
LFE20 = read.csv("~/STAA57-Project/datasets/Labour force estimates by education-2020.csv")
LFEL = read.csv("~/STAA57-Project/datasets/Labour force estimates by educational level.csv")
LFEI = read.csv("~/STAA57-Project/datasets/Labour force estimates by immigration.csv")
NOC = read.csv("~/STAA57-Project/datasets/National Occupation Classification (NOC) skill level.csv") 
WOE = read.csv("~/STAA57-Project/datasets/Wages of employees.csv")
LFEFP = read.csv("~/STAA57-Project/datasets/Labour force estimates for full and part time students.csv")

```